---
title: "CONTRIBUTING_WITH_GIT2R.Rmd"
author: "Zhian N. Kamvar"
date: "December 16, 2015"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## Motivation

The **popgenInfo** project was designed to be an educational tool that can move
at the same pace as method development by allowing people skilled in population
genetics analysis to submit workflows that demonstrate the strengths of any
given analysis in R. We use [git](https://git-scm.com/) and
[GitHub](https://github.com) to accept contributions via pull requests using
[GitHub flow](https://guides.github.com/introduction/flow/).

The basic steps of this process are:

1. **Fork** the [popgenInfo repository](https://github.com/NESCent/popgenInfo)
2. **Clone** your fork to your local machine.
3. Create a new **branch** and **commit** new changes or content.
4. **Push** your changes to your fork.
5. Open a **pull request** from your fork to NESCent.

This particular tutorial will show you the basics of contributing to the
popgenInfo website using the R package
[*git2r*](https://cran.r-project.org/package=git2r), which was developed by the
[rOpenSci project](https://ropensci.org/). This package allows for a nearly
seamless and uniform interface for all contributors regardless of operating
system.

We assume that you have already done a few things:

1. Signed up for a [GitHub account](https://github.com)
2. Install [R](https://cran.r-project.org)
3. Install [Rstudio](https://rstudio.com/desktop/download)

If you have also installed [git](https://git-scm.com/), you can set up Rstudio
to work with git to help you commit changes, pull, and push. For the purposes of
this tutorial, these things are not necessary.

# Session Information

This shows us useful information for reproducibility. Of particular importance
are the versions of R and the packages used to create this workflow. It is
considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```