---
title: "Contributing with git2r"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## Background/Motivation

The **popgenInfo** project was designed to be an educational tool that can move
at the same pace as method development by allowing people skilled in population
genetics analysis to submit workflows that demonstrate the strengths of any
given analysis in R. We use [git](https://git-scm.com/) and
[GitHub](https://github.com) to accept contributions via pull requests using
[GitHub flow](https://guides.github.com/introduction/flow/).

> **What is git?**
> 
> Git is a *version control* program. It allows you to keep track of all changes
that happen when you are writing, analyzing data, or coding. You can think of it
as a sort of electronic lab notebook. Just like you would record any weights or
measures carefully in your bench lab notebook, so should you record any analyses
you do on the computer. With git, you "add" and "commit" changes that you make.
These are analogous to writing down the values of your measurements and noting 
how you took them. 

The basic steps of the process are:

i. **Fork** the [popgenInfo repository](https://github.com/NESCent/popgenInfo)\*    
ii. **Clone** your fork to your local machine.\*

1. Checkout your master branch, **fetch** the changes from NESCent and **merge**
   them into your fork\^.
2. Create a new **branch** and then **add** and **commit** new changes or content.
3. **Push** your changes to your fork.
4. Open a **pull request** from your fork to NESCent.

\* These should only be done once.    
\^ Start here if you already have cloned your fork.

This particular tutorial will show you the basics of contributing to the
popgenInfo website using the R package
[*git2r*](https://cran.r-project.org/package=git2r), which was developed by the
[rOpenSci project](https://ropensci.org/). This package allows for a nearly
seamless and uniform interface for all contributors regardless of operating
system. A companion set of videos to go along with this tutorial can be found at
https://www.youtube.com/playlist?list=PLSFzyC3wp8-doiOcjDLlryIojK29qX2NL.

## Requirements

We assume that you have already done a few things:

1. Sign up for a [GitHub account](https://github.com)
2. Install [R](https://cran.r-project.org)
3. Install [Rstudio](https://rstudio.com/desktop/download)

If you have also installed [git](https://git-scm.com/), you can set up Rstudio
to work with git to help you commit changes, pull, and push. For the purposes of
this tutorial, these things are not necessary.

# Step i: Forking the repository (DO ONCE)

> If you have already forked and cloned the repository, go to Step 1.

Go to 
[https://github.com/NESCent/popgenInfo](https://github.com/NESCent/popgenInfo) 
and [click on the "Fork" 
button](https://help.github.com/articles/fork-a-repo/#fork-an-example-repository).
This will create a copy of the NESCent popgenInfo repository to your account.
After you've forked the repository, you never have to fork it again!

# Step ii: Clone your fork to your computer (DO ONCE)

Now that you have your repository forked to your account, you will need to 
**clone** it to your computer. Just like forking, you only have to clone your
fork to your computer once. It will live there until you decide to remove it. We
will need two things to clone our repository:

1. a place you want to store this repostiory (eg. a folder called `forks` in your `Documents`)
2. git2r

```{r, tmpdir_setup, echo = FALSE}
fork_dir <- tempdir()
```

## Create your forks folder

First, we need to make sure that we are working in the folder where we want to
set up our fork. In this tutorial, we will use `setwd()` to do this via the R
console, but it is possible to do this via Rstudio.

If you are using Windows:

```{r, eval = FALSE}
fork_dir <- "~/forks"
```

If you are using OSX:

```{r, eval = FALSE}
fork_dir <- "~/Documents/forks"
```

Now, we create the directory and then move there:

```{r, eval = FALSE}
dir.create(fork_dir)
```

We can use the function `list.files()` to see what's inside the directory:

```{r showdir, eval = FALSE}
list.files()
```

```{r showshadow, echo = FALSE}
list.files(fork_dir)
```

Currently, there's nothing here. We will use the `clone()` function from *git2r*
to make a copy of our repository. If you type `help("clone", "git2r")` in your
R console, you can see documentation about the function. We need two things
from this function:

1. The URL of the git repository
2. The path to our folder where we want to store it

The URL for the git repository is simply just the URL for your fork with `.git` 
at the end of it. For mine, it's `https://github.com/zkamvar/popgenInfo.git`. To
keep things simple, we will name the folder we want to put the repository in
"popgenInfo".

```{r clone, eval = FALSE}
library("git2r")
clone(url = "https://github.com/zkamvar/popgenInfo.git",
      local_path = "popgenInfo")
```

```{r shadow_clone, echo = FALSE}
library("git2r")
clone(url = "https://github.com/zkamvar/popgenInfo.git",
      local_path = paste0(fork_dir, "/popgenInfo"))
```


Now when we look at our directory, we can see that we have one folder called 
"popgenInfo":

```{r show_clone, eval = FALSE}
list.files()
```

```{r show_shadow, echo = FALSE}
list.files(fork_dir)
```


We can move into the folder with `setwd()` and look at all the files in there.

```{r go_clone, eval = FALSE}
setwd("popgenInfo")
list.files()
```

```{r go_shadow, echo = FALSE}
fork_pgi <- paste0(fork_dir, "/popgenInfo")
list.files(fork_pgi)
```

Now we have successfully cloned our repository into our computer using *git2r*. 
Next, we will set up our credentials and keep our master branch up to date.

## Setting up your clone

To access the repository, all you have to do is open it by double-clicking on
the "popgenInfo.Rproj" file. This will open Rstudio to this folder and set it as
the working directory. After that, in your R console, type:

```{r repository_show, eval = FALSE}
library("git2r") # in case you're starting from here
repo <- repository()
repo
```

```{r repository_shadow, echo = FALSE}
library("git2r") # in case you're starting from here
repo <- repository(fork_pgi)
repo
```

This tells R that you have a repository in this folder and that you are
referring to it as `repo`. The output of `repo` is a summary of your repository:

 - **Local** This is the branch (master) and the path where the copy of your repository is.
 - **Remote** This shows you the branch and the url of the remote repository
 - **Head** This is a summary of the last *commit* containing the date and a description.

In order to be able to synch your clone with your fork, GitHub needs to know 
that you are really who you say you are. This means that you need to associate
your name, email, and a secret token with this repository. The first two items
are easily done within R. You can use the function `config()` to set and confirm
your name and email:

```{r, config}
config(repo, 
       user.name = "Zhian Kamvar",
       user.email = "kamvarz@science.oregonstate.edu")
config(repo)
```

Next, we need to set our secret token. Known as a [**Personal Access Token
(PAT)**](https://help.github.com/articles/creating-an-access-token-for-command-line-use/),
this is a 40 character cryptographic token that is like a long and complicated
password. They both allow access to your GitHub repository, but the difference
between a PAT and a password is that a PAT can be stored as text on your
computer (which you should NEVER do with passwords) because it is easy to
generate and (most importantly) easy to remove. We can create a PAT from GitHub
and then we will store this in a file called `.Renviron` that is read every time
you restart your R session, allowing you to have this token whenever you need to
push changes to your fork on GitHub. [The instructions to generate your PAT are
here](https://help.github.com/articles/creating-an-access-token-for-command-line-use/).

Here's an example of what a `.Renviron` file with a PAT will look like on the
inside:

```
GITHUB_PAT="97cc6bf86c31a42fca2de32884cd1f1c4b1102ba"
```

Once you have created your PAT, make a text file called `.Renviron` inside of
the repository and paste your PAT in that file as shown above. This will not be
tracked by git, and will only exist on your computer.

```{r renviron_shadow, echo = FALSE}
cat(paste0("GITHUB_PAT=\"¯\\_(ツ)_/¯\""), file = paste0(fork_pgi, "/.Renviron"))
```

The next time you open R from `popgenInfo.Rproj`, this token will be available
for GitHub to ensure that you are who you say you are. It is important to note
that, if you change or delete that PAT, you must replace it here.

In the next step, we will ensure that our fork is up to date with NESCent. In 
order to do that, we need to tell our local copy that it can take updates from 
NESCent. We do this by adding NESCent as a **remote** called 'upstream' using 
the function `remote_add()` and checking to make sure it worked with the 
function `remotes()`.

```{r remote_add}
remote_add(repo, 
           name = "upstream", 
           url = "https://github.com/NESCent/popgenInfo.git")

# Check our remotes
remotes(repo)
```

With all of this information set, we can begin working on our repository.

# Step 1: Keeping your fork up to date

The first thing we need to do is make sure that our master branch is up to date.
Since other people are also contributing to this project, it is possible for
changes to be made even within seconds of you forking the repository. In this
section we will show you how you can keep your master branch up to date in your
local repository and your fork on github. 

Let's assume that you've just opened a fresh Rstudio session with the
`popgenInfo.Rproj` file. First, let's load *git2r* and access our repository.

```{r git2r_show, eval = FALSE}
library("git2r")
repo <- repository()
```

Now that we have our repository loaded, we should make sure that we are on the
**master branch**. We can do this using the `checkout()` function from *git2r*.

```{r checkout_master}
checkout(repo, "master")
repo
```

Now, whether or not we suspect changes from NESCent, it's always good form to 
update your local repository before creating any new branches. We can do it this
by first **fetching** ALL changes that have happened on the NESCent popgenInfo
repository, and then **merging** the NESCent master branch into our master
branch. We will use the *git2r* functions `fetch()` and `merge()` to accomplish
this.

First, we fetch all of the changes that have happened. If there are no changes, 
nothing will appear, but if there is new content on the master branch or new 
branches on the repository, they will be downloaded and each branch will be 
printed to your screen. For our purposes, we are only concerned with the changes
on the **master** branch, which will look something like this: 
`## [new]    b62a78c1fcd384a3ea23 refs/remotes/upstream/master`.

```{r update_master_fetch}
fetch(repo, name = "upstream")
```

It's important to note that our local repository is still unchanged at this
point as we have just downloaded the updates from NESCent. To update our local
repository, we should merge the changes from NESCent's master branch into our
master branch:

```{r update_master_merge}
merge(repo, "upstream/master")
repo
```

If there were any changes, the **Head** field will have a more current commit
than when we intialized it above.

Since we will never work directly on the master branch, there should be no 
conflicts from the NESCent branch and we now know that our branch is up to date.
Assuming everything worked, we can push this to our fork on GitHub. We will use 
*git2r*'s `push()` function to do this. We also need to supply our PAT in order
for this to work. To do so, we will use the *git2r* function `cred_token()`:

```{r push_repo, eval = FALSE}
cred <- cred_token()
push(repo, credentials = cred)
```

Now, when you check your fork, it should be updated with the most recent
changes. From here, we can checkout a new branch and then proceed with adding 
our changes or new workflow.

# Step 2: Creating a New Branch

When you add contributions, the best practice is to create a new branch off of
the master. A **branch** is a term that indicates a sort of sandbox for a
repository that can become permanent. By default, git repositories all have a
"master" branch. When you want to experiment with something else, add new
content, or simply just fix a typo, but you don't want to disturb the original
copy, you create a new branch. Good practice is to name the branch in a way that
succinctly describes what you are doing with the branch. In this step, we will
create a branch, push it to our fork, and track it.

## SETUP

Most importantly we need to make sure that we are on the master branch and it is
up to date. To double plus make sure that you are on the master branch, check it
out:

```{r chchcheckitout}
checkout(repo, "master")
repo
```

For reasons that will reveal themselves later, please run these function
definitions:

```{r fun}
my_branch <- function(x) paste0("refs/heads/", x)
my_origin <- function(x) paste0("origin/", x)
```

And, of course, don't forget your credentials:

```{r cred}
cred <- cred_token()
```


## Creation

Assuming you've updated your master branch as shown above, we can create a 
branch from the master using the function `checkout()`. Let's assume that we 
want to create a vignette for analyzing spatial statistics. I'm going to lay out
a few steps here all at once because these are the steps you want to take when
you create a brand new branch and make sure it exists on your computer and on
your fork:

1. save the branch name as a variable
2. checkout the branch
3. push the branch to the fork
4. set your local repository to track the fork

In the first step, we want to name the branch. Since we are creating a new
vignette, it's ideal for the branch name to be the same as the vignette. As
shown in the [Best Practices](CONTRIBUTING.html#best-practices) guidelines, you
should name this vignette with the date and the subject. Since we are
contributing a vignette on spatial stats and committing for the first time on
2015-12-16, the new branch and vignette will be called "2015-12-16-spatial-
stats".

In code, the steps above would look like this:

```{r checkout_branch_flow, eval = FALSE}
BRANCH <- "2015-12-16-spatial-stats"                                         # 1
checkout(repo, BRANCH, create = TRUE)                                        # 2
push(repo, name = "origin", refspec = my_branch(BRANCH), credentials = cred) # 3
branch_set_upstream(head(repo), my_origin(BRANCH))                           # 4
```

Let's examine these one by one. First, you create a variable that will hold the
name of the branch you want to create. We are using this convention to allow us
to use this name multiple times. After that, we create that branch from our
master branch. We can take a look at what our branch looks like at that point.

```{r checkout_branch, eval = FALSE}
BRANCH <- "2015-12-16-spatial-stats"  # 1
checkout(repo, BRANCH, create = TRUE) # 2
repo
```
```{r checkout_branch_shadow, echo = FALSE}
BRANCH <- "2015-12-16-spatial-stats"
# Thu Dec 17 09:57:31 2015 ------------------------------
# I have this branch created on my fork, but if I check it out normally, it will
# automatically track the fork. Instead, I am creating this branch from the
# nescent head so that it appears seamless when the user does this.
branch_create(commits(repo)[[1]], BRANCH)
checkout(repo, BRANCH, create = TRUE)
repo
branch_set_upstream(head(repo), my_origin(BRANCH))
```

Notice here that you now only have two lines in the output, **Local** and
**Head**. This is because the branch doesn't exist on your GitHub fork. This is
where the next two steps come in. We will push that branch and then make sure
that our branch is tracking the copy on the fork.

```{r push_branch, eval = FALSE}
push(repo, name = "origin", refspec = my_branch(BRANCH), credentials = cred) # 3
branch_set_upstream(head(repo), my_origin(BRANCH))                           # 4
repo
```
```{r push_branch_shadow, echo = FALSE}
repo
```

We can now see from the output that we have a **Remote** set up. Once we have 
this, we can start making changes! If you are adding a new vignette, please copy
[`TEMPLATE.Rmd`](https://github.com/nescent/popgenInfo/tree/master/TEMPLATE.Rmd)
to the **use/** directory and give it a new name that is meaningful to your 
contribution (good practice is to name it the same as your branch). If you are
including any data, place it in the **data/** directory. Once you add these
files, you need to to **add** these files and then you should **commit**, which
we will show you how to do below.

# Step 3: Adding content, committing, and pushing

If you are using Rstudio, you can use it to integrate with git. This allows you 
to do things like commit, push, and pull. Hadley Wickham, chief scientist at 
Rstudio has put up [this helpful tutorial on using git through 
Rstudio](http://r-pkgs.had.co.nz/git.html#git-status). If you aren't using
Rstudio, but you don't want to access the terminal (command line), this section
will show you basic git commands with *git2r* that will allow you to work on
your vignette and keep it up to date.

Recall that using git is analogous to keeping a good, detailed lab notebook.
When you make any changes, you record those changes (`git add`) and then say
*why* you made the changes (`git commit`).

Let's say you've already copied the `TEMPLATE.Rmd` file to the **use/**
directory and renamed it to `2015-12-16-spatial-stats.Rmd` and added some data
called `spatial_data.csv`. When you place a file in your repository, git will
not pay attention to any changes you do until you specifically add it. Until
then it is "Untracked". You can see this by running `status()`:

```{r add_stuff_shadow, echo = FALSE}
x <- file.copy(paste0(fork_pgi, "/TEMPLATE.Rmd"), 
               paste0(fork_pgi, "/use/2015-12-16-spatial-stats.Rmd"))
x <- file.create(paste0(fork_pgi, "/data/spatial_data.csv"))
set.seed(999)
m <- matrix(rnorm(20), nrow = 5, ncol = 2, dimnames = list(NULL, c("x", "y")))
write.table(m, file = paste0(fork_pgi, "/data/spatial_data.csv"), sep = ",", 
            row.names = FALSE, col.names = TRUE)
```

```{r status}
status(repo)
```

Since we've just added the files, we can use `add()` to tell git we want to
stage them for committing.

```{r add}
add(repo, "*") # adding all new or changed files
status(repo)
```

When we have done this, we can commit to our changes by using `commit()`. A
commit is basically placing a record of what you did and, importantly, why you
did it. Your commit message should record this. Commit messages should be able
to fit on a single line. Optionally, if you want to be able to add more detail,
you can enter more lines below the message:

```{r commit}
msg <- " started spatial stats vignette and added data

I copied the template to the use folder and placed
spatial data in the data folder.
"
commit(repo, msg, session = TRUE)
status(repo) # should return nothing
repo
```
> Note that we put `session = TRUE` in the commit function. This automatically
puts your R session information in the commit, which gives more information
about package versions you were working with when you were editing the vignette.

You can see a summary of what you just did by looking at the summary of your
commit:

```{r summarycommit}
summary(commits(repo)[[1]])
```


Now you can see that your commit message was recorded in the repository. Since
we see that there is a **Remote** associated with this fork, we can simply use
`push()` to push our changes up to the fork.

```{r pushit, eval = FALSE}
push(repo, credentials = cred_token())
```

# Step 4: Pull Request and Peer Review

Now that you've successfully set up git, created a new branch, and created a new
vignette, it's time to fill it with content. You can work in your vignette like
you would write in any other Rmarkdown document and keep track of your changes
like we showed above by **adding** and **committing** your changes followed by
**pushing** these up to your fork. Once you are finished with all of your
changes, you can create a pull request from your GitHub fork to NESCent.

## Creating a pull request

### Through R

You can tell R to open a browser for you to create your pull request. The URL 
for this is in the form of 
`https://github.com/USER/popgenInfo/pull/new/BRANCH`. We can create this 
URL by using the function `paste()` to glue the pieces together and then use the
`utils::browseURL()` function to open up a window. Note we are using the
variable called `BRANCH` that we named above.


```{r make_pull}
(my_account <- dirname(remote_url(repo, "origin")))
(pull_request_url <- paste(my_account, "popgenInfo/pull/new", BRANCH, sep = "/"))
```

Now that you have the URL, you can browse to it.

```{r browse_pull, eval = FALSE}
utils::browseURL(pull_request_url)
```

### Manually

If you don't want to do this from R, you can do it through your web browser.

1. Navigate to your fork on GitHub
2. Click on the dropdown menu on the left that says **"Branch: master"** and
   switch to the branch you want to create the pull request from.
3. Click on the green button that says **"New Pull Request"**.

## Peer Review

Your pull request begins a process of open peer review where content, accuracy,
and style are considered. If changes are suggested, you should revise them by
making changes on your fork and repeating the process in **Step 3**. Your pull
request will be *automatically updated* once you push the changes to your fork.

Finally, when all questions and concerns have been addressed, the pull request
may be merged into the NESCent repository as long as two of the maintainers
have signaled their approval.


# Conclusions

From this tutorial, you should have learned how to:

 - fork a repository on GitHub
 - clone that fork
 - set up credentials for *git2r*
 - update your master branch from NESCent
 - create a new branch
 - add and commit changes
 - push changes
 - create a pull request

The skills demonstrated in this tutorial are not exclusive for **popgenInfo**,
but they can be used when writing up your own reports or analyses. Using git may
not feel comfortable at this moment, but just like it takes practice to learn
how to keep a lab notebook, with practice, you will become comfortable using git
to keep track of your workflows, making your science more reproducible.

# What's Next?

Once your pull request passes peer review and is published, you can update your 
fork by starting from **Step 1**: checkout your master branch, fetch the changes
from NESCent, merge those changes, and push them up to the master branch on your
fork to make it even with NESCent.

# Contributors

 - Zhian N. Kamvar (Author)
 - Hilmar Lapp (Reviewer)
 - Margarita M. Lopéz (Reviewer)
 - Stéphanie Manel (Reviewer)

# Session Information

This shows us useful information for reproducibility. Of particular importance
are the versions of R and the packages used to create this workflow. It is
considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```
