---
title: "Template Workflow"
author: "Author name"
date: "date"
output: html_document
---

Introduction
=========
Explain the purpose of the vignette.  
Give a brief explanation of the questions that will be answered and some background information.

Assumptions
============
- Assumption 1
- Assumption 2

Data
==========
Links to information about the data e.g. package vignettes.    
Link to the data to download for the workflow.  
Format data needs to be (import data and show a section to demonstrate if needed).  

Resources/Packages required
=============
Packages required and links to packages.  

Loading the required packages:  
```{r, packages, message=FALSE}
library(hierfstat)
library(adegenet)
```

Analysis divided into sections if required
==========================================
##Example Sections
##Section 1: Load the data
```{r, loaddata, eval=FALSE}
nancycats <- import2genind("nancycats.dat")
head(nancycats)
```

##Section 2: Exploratory Data analysis/Checking assumptions  
  
##Section 3: Summary statistics  
  
##Section N: Conclusions drawn from the analysis  

Whats next
=======================
Information on further analysis that could be done, other workflows 

###Session Information

```{r,sessioninfo, echo=FALSE}
sessionInfo()




