---
title: "Template Vignette"
---

# Introduction

Explain the purpose of the vignette.  
Give a brief explanation of the questions that will be answered and some background information.
These vignettes are written in [Rmarkdown](http://rmarkdown.rstudio.com). You can find [details about markdown here](http://rmarkdown.rstudio.com/authoring_basics.html) and [details about code chunks here](http://rmarkdown.rstudio.com/authoring_rcodechunks.html).

# Assumptions

All biological analyses have assumptions.

- Assumption 1
- Assumption 2

# Data

Links to information about the data if needed e.g. package vignettes.
You can use an [external link](https://github.com/cran/adegenet/blob/master/inst/files/nancycats.gtx)
 or you can place the data [in the `data/` folder](https://github.com/NESCent/popgenInfo/tree/master/data/nancycats.gtx) (keep this < 200kb).

# Resources/Packages required

Packages required. Say where these can be found, can link to the list of packages on our repository page here also.

Loading the required packages:  

```{r, packages, message=FALSE}
library(hierfstat)
library(adegenet)
```

# Analysis divided into sections  

Example Sections

## Section 1: Load the data


```{r, load_data_show, eval=FALSE}
nancycats <- import2genind("nancycats.gtx")
summary(nancycats)
```

```{r, load_data_evaluate, echo=FALSE}
# Note: the code chunk above will be shown to the reader, but it will not run.
# This code chunk will run, but will not be shown to the reader.
nancycats <- import2genind("../data/nancycats.gtx")
summary(nancycats)
```

## Section 2: Exploratory Data analysis/Checking assumptions  
  
## Section 3: Summary statistics  
  
# Conclusions

## What's next

Information on further analysis that could be done, other workflows [can be linked as well](DifferentiationSNP.html) (note the HTML as opposed to Rmd link).

# Contributors

- Author 1
- Author 2
- Contributor 1 (role)

# Session Information

This shows us useful information for reproducibility. Of particular importance
are the versions of R and the packages used to create this workflow. It is
considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```



