---
title: "Template Vignette"
---

# Introduction

Explain the purpose of the vignette.  
Give a brief explanation of the questions that will be answered and some background information.

# Assumptions

All biological analyses have assumptions.

- Assumption 1
- Assumption 2

Data
==========
Links to information about the data if needed e.g. package vignettes.    
Link to the data to download for the workflow if it is public data or in this repository. Can download into R here and show the code.   
Format the users data needs to be in for this to work (import data and show a section to demonstrate if needed).  

Resources/Packages required
=============
Packages required. Say where these can be found, can link to the list of packages on our repository page here also.   

Loading the required packages:  
```{r, packages, message=FALSE}
library(hierfstat)
library(adegenet)
```

Analysis divided into sections  
==========================================
Example Sections
##Section 1: Load the data
```{r, loaddata, eval=FALSE}
nancycats <- import2genind("nancycats.dat")
head(nancycats)
```

##Section 2: Exploratory Data analysis/Checking assumptions  
  
##Section 3: Summary statistics  
  
##Section N: Conclusions drawn from the analysis  

Whats next
=======================
Information on further analysis that could be done, other workflows 

###Session Information

```{r,sessioninfo, echo=FALSE}
sessionInfo()




