---
title: "Creating a new R package for population genetics"
---

Motivation
=========

R packages are, at their core, standardized collections of functions, data sets,
and/or data formats. The advantage to packages is that they can be used across
all platforms, meaning that your analyses or visualizations can be easily shared
among collaborators (including your future self). If you're reading this, then
there is a good chance that you have written code that performs an analysis,
does visualization, or helps you manage data. It is very likely that these tools
you have authored will be of benefit to the larger population genetics
community. This short vignette will point you to the tools you will need in
order to get started on writing an R package for the population genetics
community and submit it to [CRAN](http://cran.r-project.org).

Tools/Resources
=========

Recently, many resources have emerged for developing R packages. We will use 
Hadley Wickham's [R packages](http://r-pkgs.had.co.nz/) for reference, but it's 
valuable to check out other tutorials.

## Tools

Writing R packages has become much easier with the advent of the IDE
[Rstudio](http://rstudio.org) and the package
[devtools](https://github.com/devtools). Using both of these in conjunction will
make developing your package much easier as they are designed to work together.

- [Download Rstudio here](http://rstudio.org/download)

You can install `devtools` by typing in your R console:

```{r, devtools_install, eval = FALSE}
install.packages("devtools", repos = "http://cran.rstudio.com/")
```

You might also need some background noise to avoid distraction. I recommend 
[the sound of a space ship control room](http://mynoise.net/NoiseMachines/spaceshipNoiseGenerator.php).

## Tutorials

- [Package Structure (Wickham)](http://r-pkgs.had.co.nz/package.html)
- [Writing an R package from scratch (cat themed) (Parker)](http://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/)
- [Developing R packages (Leek)](https://github.com/jtleek/rpackages)
- [Creating Packages (Leisch)](http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf)
- [Writing R Extensions (CRAN)](http://cran.r-project.org/doc/manuals/r-release/R-exts.html)

Creating your new package
====================

## What is in an R package?

There are four files/folders needed to create an R package:

- `DESCRIPTION`
- `R/`
- `NAMESPACE`
- `man/`

These are detailed below.

### `DESCRIPTION` 

This file simply lists all the metadata needed for the package. It lists things
such as Authors, Packages it needs to run, etc. You can view any package's
DESCRIPTION file by typing `packageDescription("package_name")`

```{r, pkgDesc}
packageDescription("poppr")
``` 
 
### `R/`

This is the folder in which all of your `.R` files live. There are different
philosophies on how to order your files in this folder. Some like to have 
[a separate R file for each function](https://github.com/dwinter/mmod/tree/master/R), while others like to
have [all functions in a single R file](https://github.com/liamrevell/Rphylip/tree/master/Rphylip/R). 
However you want to organize the files in this directory is up to you.

### `NAMESPACE`

This file lists the functions exported by your package (eg `export("myFun")`)
and functions your package imports from other packages (eg `importFrom("pegas",
"amova")`). With the modern tools available, you should not need to touch this
file.

### `man\`

All R packages need documentation. This folder contains all of the documents
that will become manual pages for your R functions. They are written in a form
similar to Latex, but completely different. Using the 
[*roxygen2*](http://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html) 
package will help immensely with writing documentation. Much like the
`NAMESPACE` file, you will not need to touch any files in this folder.

### Others

There are other folders/files that you can include, but they are completely
optional. Files like `NEWS` or `ChangeLog` give updates to changes in your
package, and folders like `inst/` and `src/` allow you to include extra data or
compiled code, respectively. See [CRAN'sdescription](http://cran.r-project.org/doc/manuals/r-release/R-exts.html#Package-structure) for details.


## Creating a package

Packages can be automatically created with the function `create()` from
*devtools*, the function `kitten()` from *pkgKitten*, or `package.skeleton()`,
which comes in default R. As a rule, I avoid things that remind me of my own
mortality, so with this tutorial, I will use `create()`.


### Setting up your information

Devtools will create a pre-set package for you, but it will be easier for you
if you set it up with your name and information, so that the DESCRIPTION file
is automatically generated with the right information. Below are the default
options for devtools. You can use `?devtools` to find out more information:

```{r, devtools_setup}
library("devtools")
options()[grep("devtools", names(options()))]
```

After looking at the documentation, we know that we want to change the values
`devtools.desc.author`, `devtools.desc.licesne`, `devtools.name`. We will also
add a "Maintainer" field to the DESCRIPTION file (tells users who to blame). The
others don't matter for now. We can add them in thusly (Note, I am using my
name; you should obviously use your own):

```{r, devtools_options}
authors_at_r <- paste0(
  "'",
  person(
    "Zhian N.", 
    "Kamvar", 
    email = "kamvarz@science.oregonstate.edu", 
    role  = c("aut", "cre")
    ),
  "'"
)

options(devtools.desc.author = authors_at_r)
options(devtools.name = "Zhian N. Kamvar")
options(devtools.desc.license = "GPL-3")
options(devtools.desc = list("Maintainer" = "'Zhian N. Kamvar' <kamvarz@science.oregonstate.edu>"))
options()[grep("devtools", names(options()))]
```

Now we can create a new package called *myFirstPackage* in our current working
directory by typing:

```{r, devtools_create, eval = FALSE}
library("devtools")
create("myFirstPackage")
```

```{r, devtools_create_real, echo = FALSE, message = FALSE}
library("devtools")
path <- file.path(tempdir(), "myFirstPackage")
create(path)
```

Now all you have to do is add in your R files and document your functions
with *roxygen2*. 

Conclusions
===========

This chapter describes the first steps to creating an R package and introduces
further resources for doing so. One of the best ways to learn how to write R
packages is to look at the source code for R packages that have already been
published on CRAN. It's easy to do this by searching the [CRAN github
repository](https://github.com/cran)

Contributors
=========

Zhian N. Kamvar
