---
title: "Individual Based Genetic Distance for SNP Data"
---

# Introduction

In this vignette, we will estimate individual genetic distances from SNP data.
It is useful when you have individual genotype data and you don't know the
populations.



# Assumptions

We will use non-evolutionary genetic distances, i.e. not based on Hardy-Weinberg
assumptions: free from assumptions.
 

# Data

The dataset used for those analysis concerns the plant: lodgepole pine (*Pinus
contorta*, *Pinaceae*). You can have more information on this data set and the
species on the web site of A. Eckert:  (http://eckertdata.blogspot.fr/). But
here the dataset is used as a test dataset with no idea of interpreting the
results in a biological way. We will work on a subset of the dataset to make
the calculations faster.
 

The data are stored in a text file (genotype = AA). We will  import the dataset
in R as a data frame, and then  convert the SNP data file into
[genind](http://www.inside-r.org/packages/cran/adegenet/docs/.valid.genind)
objects.

The dataset ["Master_Pinus_data_genotype.txt" can be downloaded 
here](https://github.com/NESCent/popgenInfo/tree/master/data/Master_Pinus_data_genotype.txt).

The  text file is a matrix of (550 rows x 3086 columns). It contains 4 extra
columns: first column is the label of the individuals, the three other are
description of the region, all the other columns are for the genotypes as (AA or
AT...).

When you import the data into R, the data file needs to be in your working
directory, or adjust the path in the `read.table()` invocation below
accordingly.


```{r, data_import_df_show, eval = FALSE}

Mydata <- read.table("Master_Pinus_data_genotype.txt", header = TRUE, check.names = FALSE)   
dim(Mydata) 
```

```{r, data_import_df_run, echo = FALSE}

Mydata <- read.table("../data/Master_Pinus_data_genotype.txt", header = TRUE, check.names = FALSE)   
dim(Mydata) 
```

```{r, data_manipulate}
ind <- as.character(Mydata$tree_id) # individual labels 
population <- as.character(Mydata$state) # population labels
county <- Mydata$county 
dim(Mydata) # 550 individuals x 3082 SNPs
```


# Resources/Packages required


Loading the required packages:  

```{r, packages, message = FALSE}
library("poppr")
library("pegas")
library("ape")
library("adegenet")
library("ade4")
```

# Analysis   

 
## Section 1: Convert the data
  

To work with the data, we need to convert the R object returned by
`read.table()` to a `genind` object. To achieve this, we create a matrix with
only genotypes, and keep only a subset of the first 100 SNP loci (to make
calculations faster). The result can then be converted to a `genind` object (for
the package adegenet). The `genind` object can then easily be converted into
`loci` objects (package pegas) (i.e. Mydata2)


```{r, data_conversion}
locus <- Mydata[, -c(1, 2, 3, 4, 105:ncol(Mydata))] 
Mydata1 <- df2genind(locus, ploidy = 2, ind.names = ind, pop = population, sep="")
Mydata1
Mydata2 <-genind2loci(Mydata1)
```





## Section 2. Individual genetic distance: euclidean distance (`dist` {adegenet})

The unit of the observation is the individuals. 

- `r nInd(Mydata1)` genotypes
- 100 binary SNPs
- Ploidy : 2 

The analysis is applied on allele frequency within individuals as represented in
the `genind` object. We can use the function `dist()` from adegenet which
provides different options. We will use the euclidean distance among vector of
allele frequencies.


```{r,Individual_dist_adegenet}

distgenEUCL <- dist(Mydata1, method = "euclidean", diag = FALSE, upper = FALSE, p = 2)
hist(distgenEUCL)

```

 
## Section 3.  Individual genetic distance: number of loci for which individuals differ (`dist.gene` {ape})  

The option `pairwise.deletion = FALSE` in the command `dist.gene()` removes all
loci with one missing values : you an see on the histogram that we get a maximum
distance of 3 loci out of 100.

We can see that we get 98 loci with at least one sample missing. Then using the
option `pairwise.deletion = TRUE` in the command `dist.gene()` allows you to keep
loci with one missing value.
 

```{r,Individual_dist_ape}

distgenDIFF <- dist.gene(Mydata2, method="pairwise", pairwise.deletion = FALSE, variance = FALSE)
hist(distgenDIFF)

# Get percent missing data per population
missing_data <- info_table(Mydata1, type = "missing")
sum(missing_data["Total", 1:100] > 0)
barplot(missing_data["Total", 1:100], xlab = "Locus", ylab = "Percent Missing")

distgenDIFF <- dist.gene(Mydata2, method="pairwise", pairwise.deletion = TRUE, variance = FALSE)
hist(distgenDIFF)

```

## Section 4: number of allelic differences between two individuals (`diss.dist` {poppr})

```{r,Individual_dist_poppr}

distgenDISS <- diss.dist(Mydata1, percent = FALSE, mat = FALSE)
hist(distgenDISS)

```



## Section 5: Conclusions drawn from the analysis  

```{r,Compare_dist}


boxplot(distgenEUCL, distgenDIFF, distgenDISS)


```

The number of allelic differences between two individuals is a different measure
from euclidean distance or number of locus differences between two individuals.

Conclusion 

### What did we learn today?

In this vignette, we explore different measures of individual genetic distances. 
It is important to investigate the different options of each command.
Missing data can be handled in different ways.


### What is next?

These individuals genetic distances can then be used in
analyses such as Mantel tests to test for isolation by distance, more complex
analyses in landscape genetics to test for resistance by distance, cluster
analysis, or spatial networks.


# Contributors

- Stéphanie Manel
- Zhian Kamvar (minor edits)

# Session Information

This shows us useful information for reproducibility. Of particular importance are the versions of R and the packages used to create this workflow. It is considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```


