---
title: "Calculating genetic differentiation from SSR data"
---

# Introduction

We will analyze a data set of bovine samples from France and West Africa using
F-statistics, AMOVA, and Discriminant Analysis of Principal Components (DAPC).
We will analyze a data set of bovine samples from France and West Africa. We
want to know if differentiation exists between Breeds or if it's based on
geographic separation.

# Data

We are going to use the `microbov` data set from the *adegenet* package. This 
data set consists of 704 bovine samples over 30 microsatellite loci (Laloe et 
al., 2007). We will also need information about the Species, Breed, and
Country/Region of origin for each sample. That data is stored in the file called
[`bovine_strata.csv` (click here to
download)](https://github.com/NESCent/popgenInfo/tree/master/data/bovine_strata.csv).

# Resources/Packages required

We are going to use *mmod* for differentiation summary statistics, *pegas* for
AMOVA, *adegenet* for DAPC, and we will use *reshape2* and *ggplot2* for
plotting our results.

Loading the required packages:  

```{r, packages, message=FALSE}
library("adegenet")
library("pegas")
library("mmod")
library("reshape2")
library("ggplot2")
```

# Analysis

## Data preparation

First, we need to load our data. Since the data is distributed with *adegenet*, 
we can load it simply by using the function `data()`. Since we want to associate
the different population strata with the data set, we also need to read in 
[`bovine_strata.csv`](https://github.com/NESCent/popgenInfo/tree/master/data/bovine_strata.csv)
as a data frame.

```{r, load_data_show, eval=FALSE}
data("microbov", package = "adegenet")
bovine_strata <- read.table("bovine_strata.csv", header = TRUE, sep = ",")
```

```{r, load_data_evaluate, echo=FALSE}
# Note: the code chunk above will be shown to the reader, but it will not run.
# This code chunk will run, but will not be shown to the reader.
data("microbov", package = "adegenet")
bovine_strata <- read.table("../data/bovine_strata.csv", header = TRUE, sep = ",")
```

The `microbov` data set is in a "genind" object. This is a way for adegenet to
keep the genetic data and deomographic data together. Let's take a look at what
our data looks like:

```{r}
microbov
head(bovine_strata) # the function head shows the first 6 rows of a data frame
```

We can see that our data matches the description above. Now we need to set
`bovine_strata` in `microbov`'s *strata* slot.

```{r, set_strata}
strata(microbov) <- bovine_strata
microbov
```

If you look at the bottom of the output, under "Optional content", you can see
that the `microbov` object now has three strata defined and we can use the
function `setPop()` to set the population from any combination of these strata.

## Summary Statistics

Perhaps the best package for assessing summary statistics for population
differentiation is [*mmod*](https://cran.r-project.org/package=mmod). This
package provides differentiation statistics presented in (Miermans and Hedrick,
2011), which scale to the
maximum value that can possibly attained, allowing for comparison across
measurements. I will not go into terrible detail about these measures or *mmod*,
but you can read more by visiting the paper linked above and viewing the [*mmod*
vignette](https://cran.r-project.org/web/packages/mmod/vignettes/mmod-demo.pdf),
off of which this section is based.

### Estimation

We will first take a look at the summary statistics for differentiation by breed
using the function `diff_stats()`:

```{r}
setPop(microbov) <- ~Breed # Use ~Country to analyze by country
diff_bov <- diff_stats(microbov)
diff_bov
```

The output in `$per.locus` are estimates of the following measures:

| Variable  | Statistic      | Details/Authority |
|-----------|----------------|-------------------|
|`Hs       `| $H_S$          | Heterozygosity with population structure |
|`Ht       `| $H_T$          | Heterozygosity without population structure |
|`Gst      `| $G_{st}$       | Nei |
|`Gprime_st`| $G_{st}^"$     | Hedrick |
|`D        `| $D$            | Jost |

The `$global` refers to global estimates of these statistics. `D_het` and
`D_mean` represent the average heterozygosity and harmonic mean, respectively.

All of these describe population differentiation, but $D$ can be interpreted
as the amount of population differentiation from 0 (panmixia) to 1 (complete
differentiation) (Hedrick and Miermans, 2011).

### Plotting

You could choose to report the global versions of these statistics, but it might
be additionally useful to show them as a boxplots with points showing their
estimates in red. I prefer to use *ggplot2* here because it allows me to easily 
overlay the observations and estimates.

```{r, plotting}
per.locus <- melt(diff_bov$per.locus, varnames = c("Locus", "Statistic"))
stats     <- c("Hs", "Ht", "Gst", "Gprime_st", "D", "D")
glob      <- data.frame(Statistic = stats, value = diff_bov$global)
head(per.locus)
head(glob)

ggplot(per.locus, aes(x = Statistic, y = value)) +
  geom_boxplot() +
  geom_point() +
  geom_point(size = rel(3), color = "red", data = glob) +
  ggtitle("Estimates of population differentiation")
```

### Confidence Intervals

We can see that we have evidence for moderate differentiation in these 
populations at the level of breed. Of course bootstrapping would be helpful to
get confidence intervals, so here we use the function `chao_bootstrap()` to
simulate new data sets and `summarize_bootstrap()` to calculate statistical
summarize based on those simulations. Since these can take a while, we're only
going to perform 10 replicates. You will want to increase that number for a more
robust analysis:

```{r bootstrap}
set.seed(20151219) # Be sure to set a seed for any random analysis!
bs_reps <- chao_bootstrap(microbov, nreps = 10)
summarise_bootstrap(bs_reps, D_Jost) # Using the D_Jost function to summarize.
```

## Analysis of Molecular Variance

AMOVA is a classical method of assessing population differentiation by 
evaluating where the most variation exists in a hierarchical population structure
(Excoffier et al., 1992). We can use this to assess population differentiation 
in countries and breeds by using the `amova()` function in *pegas*. All we need
are:

1. a distance matrix among samples
2. a data frame with hierarchical population strata

> Note: we are using the syntax `pegas::amova()` because there is another
> function called `amova()` in the *ade4* package.

```{r pegas_amova}
bov_dist  <- dist(microbov) # Euclidean distance
bov_stra  <- strata(microbov)
bov_amova <- pegas::amova(bov_dist ~ Country/Breed, data = bov_stra, nperm = 0)
bov_amova
```
If the population was panmictic, we would expect to see extremely small variance
components for Country and Breed compared to Error (i.e. the variation from
individuals within populations). We don't see that and further suspect moderate
differentiation. To test for this, you can boostrap replicates by setting
the `nperm` argument. Again, because these bootstraps may take a long time, we'll
set it to 10:

```{r pegas_amova_boot}
set.seed(20151219)
bov_amova <- pegas::amova(bov_dist ~ Country/Breed, data = bov_stra, nperm = 10)
bov_amova
```

Now we have evidence that these are significantly different.

### Other implementations

It is possible to do the same analyses in different packages, which I will list
here.

The *vegan* package has the function `adonis()` that implements a multivariate
AMOVA and takes the same kind of arguments as the *pegas* implementation of
`amova()`:

```{r adonis, eval = FALSE}
library("vegan")
set.seed(20151219)
res <- adonis(bov_dist ~ Country/Breed, data = bov_stra, permutations = 99)
```

The package *ade4* implements the original model, but it requires an abundance 
matrix of haplotypes per population. The advantage of this version is the fact 
that it calculates $\phi$ statistics for you. The package *poppr* provides a 
wrapper for the *ade4* version that will sanitize the data and allows for
calculation of within sample variance.

```{r, poppr.amova, eval = FALSE}
library("poppr")
res <- poppr.amova(microbov, ~Country/Breed, within = FALSE)
set.seed(20151219)
randtest(res, nrepet = 99)
```

## Discriminant Analyses of Principal Components (DAPC)

Finally, we will use DAPC in order to assess which populations have significant 
structure (Jombart et al., 2011). For this, we will use the function `dapc()`
from *adegenet*. 

> Warning: This analyis is sensitive to the number of principal components you
choose, so by default, you should run this function with just the data set and
without any options beforehand (`dapc(microbov)`).

```{r dapc}
setPop(microbov) <- ~Country/Breed
bov_dapc <- dapc(microbov, n.pca = 100, n.da = 4) # 100 principal components; 4 discriminant functions
scatter(bov_dapc)
```

From this output, we can see that there is a lot of differentiation based on
geography, but also that African breeds are a lot more differentiated than
European breeds.

# Conclusions

We have assessed differentiation of these cattle breeds using tools in R. From 
the diversity statistics, we can clearly see that there is moderate 
differentiation among breeds. This is supported and further bolstered by the 
results of AMOVA, showing us that there is significant differentiation among 
countries and breeds. The DAPC analysis gives us a clear picture showing highly 
differentiated African breeds with low diversity in the French breeds. Further
steps could be to assess factors such as inbreeding coefficients.

# Contributors

- Zhian N. Kamvar (Author)
- Simone Coughlan (Editor)


# References

Laloe D., Jombart T., Dufour A-B., Moazami-Goudarzi K. 2007. Consensus genetic
structuring and typological value of markers using multiple co-inertia analysis.
Genetics Selection Evolution 39:545–567. Available at:
http://dx.doi.org/10.1051/gse:2007021

Meirmans, P.G. and Hedrick, P.W., 2011. Assessing population structure: FST and
related measures. Molecular Ecology Resources, 11:5-18. Available at:
http://dx.doi.org/10.1111/j.1755-0998.2010.02927.x

Excoffier L., Smouse PE., Quattro JM. 1992. Analysis of molecular variance
inferred from metric distances among dNA haplotypes: Application to human
mitochondrial dNA restriction data. Genetics 131:479–491. Available at:
http://www.genetics.org/content/131/2/479.abstract

Jombart T., Devillard S., Balloux F. 2010. Discriminant analysis of principal
components: A new method for the analysis of genetically structured populations.
BMC genetics 11:94. Available at: http://www.biomedcentral.com/1471-2156/11/94

# Session Information

This shows us useful information for reproducibility. Of particular importance
are the versions of R and the packages used to create this workflow. It is
considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```
