---
title: "Calculating genetic differentiation and clustering methods from SSR data"
---

# Introduction

This workflow will show you how to assess population differentiation with
microsatellite/SSR data. We will analyse F-statistics, AMOVA, and Discriminant
Analysis of Principal Components (DAPC). We will analyze a data set of bovine 
samples from France and West Africa. We want to know if differentiation exists 
between Breeds or if it's based on geographic separation.

# Data

We are going to use the `microbov` data set from the *adegenet* package. This 
data set consists of 704 bovine samples over 30 microsatellite loci (Laloe et 
al., 2007; http://dx.doi.org/10.1051/gse:2007021). This data set also contains 
information about the Species, Breed, and Country/region of origin for each
sample. These data are stored in the file called [`bovine_strata.csv` (click
here to download)](https://github.com/NESCent/popgenInfo/tree/master/data/bovine_strata.csv).

# Resources/Packages required

We are going to use *mmod* for differentiation summary statistics, *pegas* for
AMOVA, *adegenet* for DAPC, and we will use *reshape2* and *ggplot2* for
plotting our results.

Loading the required packages:  

```{r, packages, message=FALSE}
library("adegenet")
library("pegas")
library("mmod")
library("reshape2")
library("ggplot2")
```

# Analysis

## Data preparation

First, we need to load our data. Since the data is distributed with *adegenet*, 
we can load it simply by using the function `data()`. Since we want to associate
the different population strata with the data set, we also need to read in
[`bovine_strata.csv`](https://github.com/NESCent/popgenInfo/tree/master/data/bovine_strata.csv)
as a data frame.

```{r, load_data_show, eval=FALSE}
data("microbov", package = "adegenet")
bovine_strata <- read.table("bovine_strata.csv", header = TRUE, sep = ",")
```

```{r, load_data_evaluate, echo=FALSE}
# Note: the code chunk above will be shown to the reader, but it will not run.
# This code chunk will run, but will not be shown to the reader.
data("microbov", package = "adegenet")
bovine_strata <- read.table("../data/bovine_strata.csv", header = TRUE, sep = ",")
```

The `microbov` data set is in a "genind" object. This is a way for adegenet to
keep the genetic data and deomographic data together. Let's take a look at what
our data looks like:

```{r}
microbov
head(bovine_strata) # the function head shows the first 6 rows of a data frame
```

We can see that our data matches the description above. Now we need to set
`bovine_strata` in `microbov`'s *strata* slot.

```{r, set_strata}
strata(microbov) <- bovine_strata
microbov
```

If you look at the bottom of the output, you can see that the `microbov` object 
now has three strata defined and we can use the function `setPop()` to set the
population from any combination of these strata.

## Summary Statistics

Perhaps the best package for assessing summary statistics for population
differentiation is [*mmod*](https://cran.r-project.org/package=mmod). This
package provides differentiation statistics presented in (Miermans and Hedrick,
2011, http://dx.doi.org/10.1111/j.1755-0998.2010.02927.x), which scale to the
maximum value that can possibly attained, allowing for comparison across
measurements. I will not go into terrible detail about these measures or *mmod*,
but you can read more by visiting the paper linked above and viewing the [*mmod*
vignette](https://cran.r-project.org/web/packages/mmod/vignettes/mmod-demo.pdf),
off of which this section is based.

We will first take a look at the summary statistics by using the function 
`diff_stats()`:

```{r}
diff_bov <- diff_stats(microbov)
diff_bov
```

The output in `$per.locus` are estimates of the following measures:

| Variable  | Satistic       | Details/Authority |
|-----------|----------------|-------------------|
|`Hs       `| $H_S$          | Heterozygosity with population structure |
|`Ht       `| $H_T$          | Heterozygosity without population structure |
|`Gst_est  `| $G_{st}$       | Nei |
|`Gprime_st`| $G_{st}^"$     | Hedrick |
|`D        `| $D$            | Jost |

The `$global` refers to global estimates of these statistics. `D_het` and
`D_mean` represent the harmonic mean and the mean, respectively.

All of these describe population differentiation, but $D$ is can be interpreted
as the amount of population differentiation from 0 (panmixia) to 1 (complete
differentiation) (Hedrick and Miermans, 2011).

You could choose to report the global versions of these statistics, but it
might be additionally useful to show them as a boxplots with points showing 
their estimates.

```{r, plotting}
per.locus <- melt(diff_bov$per.locus, varnames = c("Locus", "Statistic"))
stats     <- c("Hs", "Ht", "Gst", "Gprime_st", "D", "D")
glob      <- data.frame(Statistic = stats, value = diff_bov$global)
head(per.locus)
head(glob)

ggplot(per.locus, aes(x = Statistic, y = value)) +
  geom_boxplot() +
  geom_point() +
  geom_point(size = rel(3), color = "red", data = glob) +
  ggtitle("Estimates of population differentiation")
```



## Section 3: Summary statistics  
  
# Conclusions

## What's next

Information on further analysis that could be done, other workflows [can be linked as well](DifferentiationSNP.html) (note the HTML as opposed to Rmd link).

# Contributors

- Zhian N. Kamvar (Author)

# Session Information

This shows us useful information for reproducibility. Of particular importance
are the versions of R and the packages used to create this workflow. It is
considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```

# References

Laloe D., Jombart T., Dufour A-B., Moazami-Goudarzi K. 2007. Consensus genetic
structuring and typological value of markers using multiple co-inertia analysis.
Genetics Selection Evolution 39:545–567. Available at:
http://dx.doi.org/10.1051/gse:2007021

Meirmans, P.G. and Hedrick, P.W., 2011. Assessing population structure: FST and related measures. Molecular Ecology Resources, 11:5-18. Available at: http://dx.doi.org/10.1111/j.1755-0998.2010.02927.x