---
title: "Calculating genetic differentiation and clustering methods from SSR data"
---

# Introduction

This workflow will show you how to assess population differentiation with
microsatellite/SSR data. We will analyse F-statistics, AMOVA, and Discriminant
Analysis of Principal Components (DAPC).

# Data

We are going to use the `microbov` data set from the *adegenet* package. This
data set consists of 704 bovine samples over 30 microsatellite loci (Laloe et
al., 2007; http://dx.doi.org/10.1051/gse:2007021). This data set also contains
information about the Species, Breed, and Country of origin for each sample.
These data are stored in the file called [`bovine_strata.csv` (click here to
download)](https://github.com/NESCent/popgenInfo/tree/master/data/bovine_strata.csv).

# Resources/Packages required

Packages required. Say where these can be found, can link to the list of packages on our repository page here also.

Loading the required packages:  

```{r, packages, message=FALSE}
library(hierfstat)
library(adegenet)
```

# Analysis divided into sections  

Example Sections

## Section 1: Load the data


```{r, load_data_show, eval=FALSE}
nancycats <- import2genind("nancycats.gtx")
summary(nancycats)
```

```{r, load_data_evaluate, echo=FALSE}
# Note: the code chunk above will be shown to the reader, but it will not run.
# This code chunk will run, but will not be shown to the reader.
nancycats <- import2genind("../data/nancycats.gtx")
summary(nancycats)
```

## Section 2: Exploratory Data analysis/Checking assumptions  
  
## Section 3: Summary statistics  
  
# Conclusions

## What's next

Information on further analysis that could be done, other workflows [can be linked as well](DifferentiationSNP.html) (note the HTML as opposed to Rmd link).

# Contributors

- Zhian N. Kamvar

# Session Information

This shows us useful information for reproducibility. Of particular importance
are the versions of R and the packages used to create this workflow. It is
considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```

# References

Laloe D., Jombart T., Dufour A-B., Moazami-Goudarzi K. 2007. Consensus genetic structuring and typological value of markers using multiple co-inertia analysis. Genetics Selection Evolution 39:545–567. Available at: http://dx.doi.org/10.1051/gse:2007021

