---
title: "Population Differentiation for Sequence Data"
author: "Margarita M. López-Uribe"
date: "March 19, 2015"
output: html_document
---
Introduction
=========
Now, we will discuss in more detail how to assess population genetic structure from sequence data. Assuming that you have a priori information about how the individuals are grouped in subpopulations, one can do the following analyses: 1) quantify pairwise subpopulation structure and their significance, 2) test for hierarchical structure among groups of subpopulations, and 3) use population clustering algorithms to corroborate the a priori grouping hypothesis. We will go into each of these analyses in this vignette.  


Resources/Packages
=========
```{r, packages, message=FALSE}
library(apex)
library(adegenet)
library(ape)
library(pegas)
library(hierfstat)
library(mmod)
library(genetics)
library(poppr)
```

Import data
========
We will first upload FASTA files for each gene independently, and then we will create a multidna object that concatenates both genes (allData1). Finally, this multidna object (allData2) will be converted into a genind object (allData3), which will be used for downstream analyses.
```{r, data_import_FASTA, eval=TRUE}
myData1<- read.FASTA("Ebom_mt.fas") 
tmp1<-as.matrix(myData1)
tmp1
myData2<- read.FASTA("Ebom_CAD.fas") 
tmp2<-as.matrix(myData2)
tmp2
allData1<-list(mt=tmp1[,1:1271], CAD=tmp2[,1:673])
tmp3<-as.matrix(allData1)
allData2<-new("multidna",allData1)
plot(allData2, cex=0.2)
allData3<-multidna2genind(allData2,mlst=TRUE)
pop(allData3)<-(rep(c("CA","Ch","Am","AF"), each=10))
allData3
```

```{r, data_import_FASTA_real, eval = FALSE}
myData1<- read.FASTA("../data/Ebom_mt.fas") # this creates a DNAbin object
tmp1<-as.matrix(myData1) # converts the list of sequences into a matrix
tmp1
myData2<- read.FASTA("../data/Ebom_CAD.fas") # this creates a DNAbin object
tmp2<-as.matrix(myData2) # converts the list of sequences into a matrix
tmp2
allData<-list(mt=tmp1[,1:1271], CAD=tmp2[,1:673]) #create a multidna object concatenating two genes: mt and CAD
pop(allData3)<-(rep(c("CA","Ch","Am","AF"), each=10)) #assign populations to each individual
allData3
```

Pairwise population differentiation
========
### Overall F-statistics
```{r, Fstats1}
diff_stats(allData3) # this function calculates overall Nei's Gst, Hedrick's Gst and  of the dataset
Phi_st_Meirmans(allData3) # this function calculates overall PhiST, the Fst analog for DNA sequence data
```
### Pairwise Fst
```{r, Fstats2}
pairwise_Gst_Nei(allData3, linearized=FALSE) #Calculates pairwise Gst. If linearized =TRUE, it calculates 1/(1-Gst)  
pairwise_Gst_Hedrick(allData3, linearized=FALSE)#Calculates pairwise Gst. If linearized =TRUE, it calculates 1/(1-Gst')  
pairwise_D(allData3, linearized=FALSE, hsht_mean="harmonic") #Calculates pairwise Gst. If linearized =TRUE, it calculates 1/(1-D)  
```
### Testing for significance
To estimate if populations are significantly different, we will generate 1000 replicates of the dataset using the function chao_bootstrap(). Then, summary statistics (meand and 95% CI) will be calculated for each of the different parameters of population differentiation. 
```{r, Fstats3}
bs<-chao_bootstrap(allData3, nreps=100)
summarise_bootstrap(bs, Gst_Nei) #for Nei's Gst
summarise_bootstrap(bs, Gst_Hedrick) #for Hedrick's Gst
summarise_bootstrap(bs, D_Jost) #for Jost's D
```

AMOVA (Analysis of Molecular Variance)
========
Analysis of Molecular Variance (AMOVA) is a method for estimating population differentiation from molecular data taking into account the mutational distance between alleles. Unlike Fst, which quantifies genetic differentiation based on allele frequencies, AMOVA treats molecular data as vectors and estimates Euclidean distances between alleles. Furthermore, it is possible to test hypotheses about differentiation by grouping subpopulations in a hierarchical structure (Ref).
```{r}
allData4<-genind2hierfstat(allData3)
allData_dist<-dist.multidna(allData2) # this calculates pairwise Eucledian distances among individuals
levels <- factor(c(rep("West", 20), rep("East", 20)))
populations <- factor(c(rep("CA", 10), rep("Ch", 10),rep("Am", 10), rep("AF", 10)))
varcomp.glob(levels=levels)
amova(allData_dist ~ populations, nperm = 100)
amova(allData_dist ~ levels, nperm = 100)
#poppr.amova(allData3, ~populations/groups)
