---
title: "Population Differentiation for Sequence Data"
author: "Margarita M. López-Uribe"
date: "March 19, 2015"
output: html_document
---
Introduction
=========
Now, we will discuss in more detail how to assess population genetic structure from sequence data. Assuming that you have a priori information about how the individuals are grouped in subpopulations, one can do the following analyses: 1) quantify pairwise subpopulation structure and their significance, 2) test for hierarchical structure among groups of subpopulations, and 3) use population clustering algorithms to corroborate the a priori grouping hypothesis. We will go into each of these analyses in this vignette.  


Resources/Packages
=========
```{r, packages, message=FALSE}
library(adegenet)
library(ape)
library(pegas)
library(hierfstat)
library(mmod)
library(genetics)
```

Import data
========
Refer to vignette on "Summary Statistics for Sequence Data" (add link) for detailed instructions on how to upload one locus or multilocus FASTA files for this type of analyses. Here is the code:
```{r, data_import_FASTA, eval=FALSE}
myData1<- read.FASTA("Ebom_mt.fas") 
tmp1<-as.matrix(myData1)
myData2<- read.FASTA("Ebom_CAD.fas") 
tmp2<-as.matrix(myData2)
allData<-cbind(tmp1,tmp2)
tmp3<-as.matrix(allData) 
allDatab<-as.genind.DNAbin(tmp3, rep(c("CA","Ch","Am","AF"), each=5)) 
```

```{r, data_import_FASTA_real, echo = FALSE}
myData1<- read.FASTA("../data/Ebom_mt.fas") # this creates a DNAbin object
tmp1<-as.matrix(myData1) # converts the list of sequences into a matrix
myData2<- read.FASTA("../data/Ebom_CAD.fas") # this creates a DNAbin object
tmp2<-as.matrix(myData2) # converts the list of sequences into a matrix
allData<-cbind(tmp1,tmp2)
tmp3<-as.matrix(allData) 
allDatab<-as.genind.DNAbin(tmp3, rep(c("CA","Ch","Am","AF"), each=5))
```

Pairwise population differentiation
========
### Overall F-statistics
```{r, Fstats1}
allData_loci<- as.loci(allDatab) 
Fst(allData_loci) # this function calculates overall Fit, Fst, and Fis of the dataset
```
### Pairwise Fst
```{r, Fstats2}
allData_loci<- as.loci(allDatab) 
Fst(allData_loci) # this function calculates overall Fit, Fst, and Fis of the dataset
```

AMOVA (Analysis of Molecular Variance)
========
Analysis of Molecular Variance (AMOVA) is a method for estimating population differentiation from molecular data taking into accound the mutational distance between alleles. Unlike Fst, which quantifies genetic differentiation based on allele frequencies, AMOVA treats molecular data as vectors and estimates Eucledian distances between alleles. Furtheremore, it is possible to test hypotheses about differentiation by grouping subpopulations in a hierarchical structure (Ref).
```{r}
allData_dist<-dist.dna(allData) # this calculates pairwise Eucledian distances among individuals
groups <- factor(c(rep("West", 10), rep("East", 10)))
populations <- factor(c(rep("CA", 5), rep("Ch", 5),rep("Am", 5), rep("AF", 5)))
amova(allData_dist ~ groups, nperm = 100)
amova(allData_dist ~ populations, nperm = 100)
```

