---
title: "Calculating Basic Population Genetic Statistics from microsatellite Data"
---

# Introduction

In this vignette, you will calculate basic population genetic statistics from microsatellite data using R packages. 

1. Genetic diversity,
2. Test Hardy Weinberg 
3. $F_{is}$ and global $F_{st}$

# Assumptions

H0: we assume that the microsatellites are in Hardy-Weinberg equilibirum
H1: they are not.



# Data

The dataset used for these analyses are a R dataset:nancycats
It contains microsatellites genotypes of 237 cats from 17 colonies of Nancy (France). Each individuals of the 17 colonies are genotyped for 9 microsatellites.  

nancycats is a genind, i.e. an object of the library adegenet

# Resources/Packages required


```{r, packages, message=FALSE}
 
library(adegenet)
library(pegas)
```

 
 

## Section 1: Load the data
After loading the 'genind' object, we convert it into a `loci` objects (package pegas)

```{r, load_data_show, eval=FALSE}
nancycats  
summary(nancycats)

```


```{r, data_conversion, eval=FALSE}
Mydata <- genind2loci(nancycats)

```


## Section 2: Genetic diversity (osberved and expected heterozygosity)   


```{r, Genetic_diversity, eval=FALSE}
nancycats  
div<-summary(nancycats)
div
plot(div$Hobs, xlab="Loci number", ylab="Observed Heterozygosity", main="Observed heterozygosity per locus")

plot(div$Hobs,div$Hexp, xlab="Hobs", ylab="Hexp", 
   main="Expected heterozygosity in function of observed heterozygosity per locus")

bartlett.test(list(div$Hexp, div$Hobs)) #  
```

We get various information from the command summary. 
The one that interest us is the observed and expected heterozygotity per locus.
We observed that heterozygosity varies among loci.

The test of bartlett indicates that we have no difference between the mean observed and expected heterozygotity

## Section 3: Testing for Hardy-Weinberg Equilibrium

```{r,Hardy-Weinberg_test}

hw.test(Mydata, B = 1000)
```


We get for each locus a test of significance of the null hypothesis: $H_0$ the
locus is in HW equilibrium in the population/ $H_1$. The locus is not in HW equilibrium.

We can conclude from the p-values of each test  that any locus is in HW equilibrium.

# Conclusions
#  What did we learn today?

In this vignette, we learned how to explore the patterns of genetic diversity  in one population.
Also, you have an idea of potential violations of the dataset to the null Wright-Fischer model.

## What is next?

You may now want to move into looking into population differentiation   (See [Calculating genetic differentiation and clustering methods from microsatellites data](DifferentiationMicrosatellites.html))

 

# Contributors

- Stéphanie Manel
 

# Session Information

Information about the R session used for this tutorial.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```



