---
title: "Calculating Basic Population Genetic Statistics from Sequence Data"
---

Introduction
=========

In this vignette, you will see general information on how to calculate basic
population genetic statistics from sequence data using R packages. These basic
statistics serve as exploratory analyses that will tell you 1) the levels of
genetic diversity in your overall population and subpopulations, 2) general
patterns of the distribution of genetic diversity across subpopulations, and 3)
possible violations of the data to the Wright-Fischer Model (Ref).

The first section of this tutorial provides information on how to analyze
sequence data from one locus (link to section 1). The second section will
discuss how to analyze multilocus sequence data (link to section 2).


Resources/Packages
=========
```{r, packages, message=FALSE}
library(adegenet) # to store the data
library(ape)
library(pegas)
library(hierfstat)
library(mmod)
library(genetics)
```

Section I - Workflow for Raw Sequence Data
========

### Import data
To import our raw sequence data into R, the most widely used formats are
[FASTA](http://www.ncbi.nlm.nih.gov/BLAST/fasta.html) and nexus. Before you
import your data into R, please make sure the sequences are aligned. Sequence
alignments can be done in software such as Clustal or Muscle. Both are are part
of the software SeaView. Alignments can be saved in FASTA format. Another R
option could be `library(Biostrings)` but it is not discussed here. In all
cases, we will convert the sequence data file into
[genind](http://www.inside-r.org/packages/cran/adegenet/docs/.valid.genind)
objects, which is the standard file format for population genetic analyses in R.

Dataset "Ebom_mt.fas" can be downloaded 
[here](https://github.com/NESCent/popgenInfo/tree/master/data/Ebom_mt.fas).    
Dataset "Ebom_CAD.fas" can be downloaded 
[here](https://github.com/NESCent/popgenInfo/tree/master/data/Ebom_CAD.fas).

Import the previously aligned sequence data into R. The 'Ebom_mt.fas' file
contains alignments of two concatenated mitochondrial genes.

```{r, data_import_FASTA_locus1, eval = FALSE}
myData1 <- read.FASTA("Ebom_mt.fas") # this creates a DNAbin object
class(myData1)
```

```{r, data_import_FASTA_locus1_real, echo = FALSE}
myData1 <- read.FASTA("../data/Ebom_mt.fas") # this creates a DNAbin object
class(myData1)
```

To convert DNAbin objects to genind objects:

```{r, data_import_DNAbin}
#library(mmod)
tmp1 <- as.matrix(myData1) # converts the list of sequences into a matrix
myData1b <- as.genind.DNAbin(tmp1, rep(c("CA","Ch","Am","AF"), each=5))
```

### Genetic diversity indices:
Calculate nucleotide diversity and summary descriptors for the data:

```{r, pi}
myData1_pi <- nuc.div(myData1) # nucleotide diversity from a sample of DNA sequences in DNAbin format
myData1_pi
myData1_summ <- summary(myData1b) # to visualize summary descriptors of your sequence data file
```
Make some plots:

```{r,plot}
par(mfrow = c(2, 2))
plot(myData1_summ$pop.eff, myData1_summ$pop.nall, xlab = "Sample size",
	 ylab = "Number of alleles", main="Alleles numbers and sample sizes",
	 type = "n")
text(myData1_summ$pop.eff, myData1_summ$pop.nall, lab = names(myData1_summ$pop.eff))

barplot(myData1_summ$loc.nall, ylab="Number of alleles", 
		main = "Number of alleles per locus")

barplot(myData1_summ$Hexp-myData1_summ$Hobs, ylab = "Hexp - Hobs",
		main = "Heterozygosity: expected-observed")
barplot(myData1_summ$pop.eff, main = "Sample sizes per population",
		ylab = "Number of genotypes", las = 3)
```

Note that no observed heterozygosity is expected because sequences come from
mitochondrial data.

### Testing for Hardy-Weinberg Equilibrium
```{r, HWE1, message=FALSE, warning=FALSE}
#library(genetics)
myData1_HWE <- HWE.test.genind(myData1b) # this will test for Hardy-Weinberg equilibrium
myData1_HWE # p-values lower than 0.05 indicate deviations from HWE
```
Note: If you have loci in your dataset that deviate from HWE, your dataset may violate assumptions for downstream population genetic analyses.

### F statistics
```{r, Fstats}
myData1_loci <- as.loci(myData1b) 
Fst(myData1_loci) # this function calculates overall Fit, Fst, and Fis of the dataset
```

### Testing for Population Differentiation

#### Testing for Population Differentiation from diploid data

```{r, Fst1}
myData1_bs <- chao_bootstrap(myData1b, nreps=100) # subsample your dataset multiple times to estimate significance of genetic differentiation statistics 
summarise_bootstrap(myData1_bs, Gst_Nei) # estimate Gst_Nei and whether or not it is significant
summarise_bootstrap(myData1_bs, Gst_Hedrick) # estimate Hedrick's Gst and whether or not it is significant
summarise_bootstrap(myData1_bs, D_Jost) # estimate Jost's D and whether or not it is significant
myData1_GenDiff <- diff_stats(myData1b) # general test for genetic differentiation per locus and across all subpopultions using Gst, Gst' and Jost D (estimated and based on the harmonic mean between loci)
myData1_GenDiff
```

#### Testing for Population Differentiation from haploid data

```{r, Fst2}
#library(mmod)
myData1_Phist <- Phi_st_Meirmans(myData1b) 
myData1_Phist # test fo population differentation for haploid data
```



Section II - Workflow for Multilocus Sequence Data 
========
### Import data

To import raw sequence data from more loci into R:

```{r, data_import_FASTA_locus2, eval=FALSE}
myData2 <- read.FASTA("Ebom_CAD.fas") # this creates a DNAbin object
class(myData2)
tmp2    <- as.matrix(myData2) # converts the list of sequences into a matrix
allData <- cbind(tmp1, tmp2)
```

```{r, data_import_FASTA_locus2_real, echo = FALSE}
myData2 <- read.FASTA("../data/Ebom_CAD.fas") # this creates a DNAbin object
tmp2 <- as.matrix(myData2) # converts the list of sequences into a matrix
allData <- cbind(tmp1,tmp2)
```

To convert DNAbin objects to genind objects:

```{r, data_import_DNAbin_all}
#library(mmod)
tmp3 <- as.matrix(allData) # converts the list of sequences into a matrix
allDatab <- as.genind.DNAbin(tmp3, rep(c("CA","Ch","Am","AF"), each=5)) # bins data from multiple loci into one DNAbin object - MAKE SURE SEQUENCES IN THE FASTA FILE HAVE THE SAME NAME
```

#Analyses

For example, testing for Genetic Differentiation:

```{r, HWE2, message=FALSE}
#library(genetics)
allData_HWE <- HWE.test.genind(allDatab, res="matrix") # this will generate a matrix of p-values for HWE
allData_HWE

```

For example, testing for overall Genetic Differentiation:

```{r, Fstats2}
allData_loci <- as.loci(allDatab) 
Fst(allData_loci) # this function calculates overall Fit, Fst, and Fis of the dataset
```

Conclusions
========

### What did we learn today?

In this vignette, we learned how to explore the patterns of genetic diversity in
our dataset. Now, you know the level of genetic differentiation in your
population and subpopulations, and whether or now they are genetically
different. Also, you have an idea of potential violations of the dataset to the
null Wright-Fischer model.

### What is next?

You may now want to move into looking into population differentiation in more
detail see [Population Differentiation for Sequence Data](PopDiffSequenceData.html). You may also want to explore demographic parameter
estimation (e.g. deviations from neutrality, effective population size).

# Contributors

- Margarita M. López-Uribe

# Session Information

This shows us useful information for reproducibility. Of particular importance are the versions of R and the packages used to create this workflow. It is considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 100)
devtools::session_info()
```
